---
title: "Modelling the Measles Outbreak of 1861 in Hagelloch"
date: "`r Sys.Date()`"
output: html_document
---

# Introduction

Data were collected from a Measles outbreak in an isolated German village, Hagelloch, in 1861. Available in the surveillance R package. Data on the 188 cases in the measles outbreak among children were originally collected by Dr. Albert Pfeilsticker at the time of the outbreak and published in a paper in 1863. The dataset was augmented and re-analysed by Dr. Heike Oesterle (1992).

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(deSolve)
library(surveillance)
library(patchwork)
library(EpiEstim)
library(glue)
library(gt)
library(gganimate)
```

# Get Hagelloch dataset
```{r get_hagelloch_data}
# image for the visualisation
png_town <- png::readPNG('Hagelloch.png')

# load the data from surveillance package
data("hagelloch")
df_measles_orig <- hagelloch.df

# get a subset of variables
df_measles <- df_measles_orig %>% 
    # drop a very late case 
    filter_out(PN == 141) %>%
    # useful variables 
    select(PN, IFTO, AGE, PRO, tPRO, ERU, tR, SI, GE, CL, DEAD, x.loc, y.loc) %>% 
    mutate(days_infected = round(tR - tPRO)) %>% 
    identity()

remove(hagelloch.df, hagelloch, df_measles_orig)

```

# create animation dataset
```{r create_animation_dataset}
# first symptom date
date_start <-  min(df_measles$PRO)
# final rash start date
date_final <- max(df_measles$ERU)

# assume exposure (average) 10 day before symptoms
# allow 5 days after last rash appears for charts to fade out
days_fade_in <- 11
days_fade_out <- 5
df_dates <- 
        # days before the first symptoms, so Exposed compartment is in this time period
        tibble(date = seq(as.Date(date_start - days_fade_in), as.Date(date_start - 1), by='day')) %>% 
        # days from first symptoms to final rash appearing, so Infectious compartment is in this period
        bind_rows(tibble(date = seq(as.Date(date_start), as.Date(date_final), by='day'))) %>% 
        # days after last rash appears, so all Infections will move into Recovered compartment during this time 
        bind_rows(tibble(date = seq(as.Date(date_final + 1), as.Date(date_final + days_fade_out), by='day'))) %>%
        # make a day number where first symptoms are on day 1 
        mutate(day_outbreak = row_number() - days_fade_in) %>% 
        mutate(day_plot = day_outbreak + days_fade_in) %>% 
        identity()

message(paste0('... first symptom ',date_start,' (day 1), final rash started ', date_final))
remove(days_fade_out, date_start, date_final)

# location. As a family shares the same location, add some noise so they show as clusters
jitter <- 4
df_animate_location <- df_measles %>% 
    # each x, y location has multiple people, create small clusters
    select(PN, x.loc, y.loc) %>% 
    mutate(x.locjit = x.loc + rnorm(n(), 0, jitter)) %>% 
    mutate(y.locjit = y.loc + rnorm(n(), 0, jitter)) %>% 
    identity()

# day of the outbreak some individuals died
df_animate_died <- df_measles %>% 
    select(PN, DEAD) %>% 
    left_join(df_dates, by=c('DEAD' = 'date')) %>% 
    mutate(day = coalesce(day_outbreak, 0)) %>% 
    select(PN, died = day_plot)

# day of the outbreak when an individual moves compartments
df_compartment_days <- df_measles %>% 
    mutate(day_I = round(tPRO)) %>% 
    mutate(day_R = round(tR) + 1) %>% 
    mutate(day_E = day_I - 10) %>% 
    select(PN, day_E, day_I, day_R )

# make a df of compartment each outbreak day for each individual 
df_animate <- 
    cross_join(df_dates, df_compartment_days) %>% 
    arrange(PN) %>% 
    mutate(compartment = case_when(day_outbreak < day_E ~ 'S',
                                day_outbreak < day_I ~ 'E',
                                day_outbreak < day_R ~ 'I',
                                day_outbreak >= day_R ~ 'R')) %>%  
    # make compartment a factor to get the order correct
    mutate(compartment = factor(compartment, levels=c('S','E','I','R'))) %>% 
    group_by(PN, compartment) %>% 
        mutate(days_in_compartment = case_when(compartment == 'S' ~ 2,
                                               compartment == 'R' ~ 3,
                                               compartment == 'E' ~ row_number(),
                                               compartment == 'I' ~ 11 - row_number())) %>% 
    ungroup() %>% 
    left_join(df_animate_location, by = 'PN') %>% 
    left_join(df_animate_died, by = 'PN') %>% 
    identity()

remove(df_animate_location, df_animate_died)
```

# generate a frame for each day and stitch together into a gif
```{r plot_functions}

count_days <- max(df_animate$day_plot)
compartment_colour <- c(S = "green3", E = "orange", I = "red", R = "grey", X = 'white')

plot_1 <- function(df_plot, plot_day, background_img) {

    # give some brief narration
    current_I <- df_plot %>% filter(day_plot == plot_day, compartment == 'I') %>% nrow()
    current_E <- df_plot %>% filter(day_plot == plot_day, compartment == 'E') %>% nrow()
    current_D <- df_plot %>% filter(!is.na(died), day_plot == plot_day, died <= plot_day) %>% nrow()
    first_mortality <- min(df_plot$died, na.rm = TRUE)

    info_text <- glue::glue('Day: {plot_day - days_fade_in} of {count_days - days_fade_in}')
    if (plot_day <= days_fade_in) {
        info_text <- glue::glue('Day: {plot_day - days_fade_in} : Infection is building before symptoms appear, exposed = {current_E}')
    } else if (plot_day >= 11 & plot_day < first_mortality) {
        info_text <- glue::glue('{info_text} : Symptoms appear, exposed = {current_E}, infected = {current_I}')
    } else if (plot_day >= first_mortality) {
        info_text <- glue::glue('{info_text} : exposed = {current_E}, infected = {current_I}, died = {current_D}')
    }

    # geographical plot
    p1 <- df_plot %>% 
        filter(day_plot == plot_day) %>% 
        ggplot(aes(x.locjit, y.locjit, 
               colour = compartment, 
               size = days_in_compartment,
               alpha = 0.3)) +

        annotation_raster(
            background_img,
            xmin = -Inf, 
            xmax = 400,
            ymin = -40, 
            ymax = Inf
        ) +

        geom_point() +

        # show deaths
        geom_text(
            data = df_plot %>% filter(died > 0 & died <= plot_day),
            aes(label = "⚠️"),
            size = 8
        ) +

        # setting : colours come from the factor levels
        scale_colour_manual(values = compartment_colour) +
        # setting : size come from the data, not automatic aes rescaling
        scale_size_identity() +
        labs(title = 'Visualising the Hagelloch Measles outbreak of 1861',
            subtitle = glue::glue('{info_text}'),
            x = 'x', 
            y = 'y') +
        theme_minimal() +
        theme(
            axis.text.x = element_blank(),
            axis.title.x = element_blank(),
            panel.grid = element_blank()
        ) +
        theme(legend.position = 'none')
}

# plot of compartment growth
plot_2 <- function(df_plot, plot_day) {

    p2 <- df_plot %>% 
            count(day_plot, compartment) %>% 
            # we need all the days to make the x axis static
            # but points ahead of this day need to be invisible
            mutate(compartment = if_else(day_plot <= plot_day, compartment, 'X')) %>% 
            mutate(compartment = factor(compartment, levels=c('S','E','I','R','X'))) %>% 
            ggplot(aes(day_plot, n, colour = compartment)) +
            geom_point() +
            scale_colour_manual(values = compartment_colour) +
            theme_minimal() +
            labs(title = 'Compartment size') +
            theme(
                axis.text.x = element_blank(),
                axis.title.x = element_blank(),
                panel.grid = element_blank()
            ) +
            theme(legend.position = 'none')
}

# we need a single horizontal line to divide the plots
plot_divider <- function() {
    divider <- ggplot() + 
               geom_hline(yintercept = 0, linewidth = 0.3, colour = 'grey') +
               theme_void()
}

```

```{r test_images}

# create images to test specific plot days
xx <- seq(9,12)
for (x in xx) {
    print(x)
    p1 <- plot_1(df_animate, x, png_town)
    print(p1)
}
```

```{r create_animation}

# create a sequence of the days for each image
days <- seq(1, count_days)
# create the divider line plot
p_divide <- plot_divider()

# create an image for each day
for (d in days) {  

    p1 <- plot_1(df_animate, d, png_town)
    p2 <- plot_2(df_animate, d)
    p <- (p1 / p_divide / p2) + plot_layout(heights = c(4, 0.1, 0.8))
    ggsave(filename = glue("png/frame_{sprintf('%03d', d)}.png"), plot = p, width = 8, height = 6, dpi = 150)

    if (d == length(days)) {
        message(glue::glue(' ... Finished. {d} frames created'))
    } else if (d %% 10 == 0) {
        message(glue::glue(' ... {d} frames created'))
    }
}

# stick them together into a gif
message(glue::glue(' ... Creating gif'))
png_files <- list.files("png", pattern = "png$", full.names = TRUE)
img <- magick::image_read(png_files) %>% magick::image_animate(fps = 0.5)
magick::image_write(img, "outbreak_animation.gif")

```
